### 2023-01-24 Manuscript/Draft1.Rmd
add a few Chris comments in methods

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock. There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity. The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates. Model-based abundance estimates are often derived from model framewroks such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects. The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance]. VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

@@

Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process **cite?** and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. **BL: Help with correct notation for adding noise**. We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Tabl XXX lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.

@@

General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors. We also see individual cases where the stratified mean produced the lowest absolut error and instances where including covarites in VAST models actually increase the absolute error. When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error [[as expected]], though there are some scenarios where the impact is minimal.

>>>>

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock. There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity. The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates. Model-based abundance estimates are often derived from model framewroks [[frameworks]] such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects. The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance]. VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and [[can?]] take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

@@

Our goal is to determine indexing approaches and settings that are robust to future environmental conditions and resulting spatial biomass patterns. An underlying assumption in all indexing methods is that individual random samples combine to accurately represent true abundance by a) containing a low enough noise level in the samples to allow for a discernible pattern and b) sampling all strata in which the population exists. These assumptions can be questioned given enough noise in the sampling process **cite?** and/or climate change causing a population to move into previously uninhabited strata. To simulate the impact of noise, we compare indexing estimates after adding noise to our samples versus those using the true sampling values. **BL: Help with correct notation for adding noise**. We simulate the effect of populations moving into new habitat by comparing indexing estimates using samples from all strata inhabited by each species on Georges bank [[Bank]] to those that only include a subset of the full spatial domain for each species. We chose strata to exclude for each species by reviewing how spatial preferences evolved in our increasing temperature scenarios and removing strata that each species either shifted into, or away from. Tabl XXX [[needs to be added]] lists all strata inhabited by each species, those that are removed from certain calculations, and the explanation of why these strata were removed.

@@

General themes that exist in Tables X, Y and Z are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors. We also see individual cases where the stratified mean produced the lowest absolut [[absolute]] error and instances where including covarites in VAST models actually increase the absolute error. When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error [[as expected]], though there are some scenarios where the impact is minimal.
